Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-01-26 doc/invt.tex
invt.tex: editing changes

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of free energy calculation methods including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the noise in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme (WL or metadynamics),
  and the initial updating magnitude should be half of
  the previous equilibrium value.
  %
  Further,
  the WL updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is the inverse number of sampling steps with a proper shift,
  as given by the $1/t$ prescription.
  %
\end{abstract}

@@

The Wang-Landau (WL) algorithm\cite{wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
Both techniques employ regular updating of the bias potential
to discourage future visits to the current configuration
by incrementally elevating the bias potential there.
%
The difference lies
in the updating window function,
which takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in the WL case,
or that of a Gaussian
in the metadynamics.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, zhou2008},
and for convergence
one  has to gradually decrease
the updating magnitude.

@@

In this study,
we present a method of computing
the optimal schedule
for a general adaptive FDS method,
including the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanic one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
We will show that the optimal schedule
depends on how the bias potential is updated,
and hence is different in the WL and metadynamics cases.
%
While the optimal schedule in the WL case
is given by the known inverse-time formula,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
In either case, however,
the initial updating magnitude
can be optimally set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time in the WL case.
%
Further, we can generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the above origin-shifted inverse-time schedule
turns out to be optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help to better
plan and optimize FDS simulations.

@@

For a large system at temperatures of interest,
the distribution $p^*_i$ often tends to
be localized,
%
and to explore the global shape
of the PMF, $-\ln p^*_i$,
it is often advantageous to perform
biased sampling that targets
a wider and smoother distribution, $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

@@

We wish the biased equilibrium distribution, $\pi_i$,
to be the same as the intended distribution, $p_i$,
which is usually flat,
i.e., $p_i$ constant for all $i$'s,
or nearly flat\cite{
dayal2004, *trebst2004, *zhang2007, barducci2008, singh2011}.
%
We will generally call such a biased simulation
a flat-distribution-sampling (FDS) simulation.
%
To fulfill this goal, the bias potential must satisfy
%
\begin{equation}
  V_i \to \ln p^*_i - \ln p_i + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
according to Eq. \eqref{eq:pi_p_V1}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
This relation thus sets a practical goal for the convergence of FDS
methods---adjusting the bias potential to make the
observed distribution, or the normalized histogram,
match the intended one.
%
Once this aim is achieved,
the PMF, $-\ln p^*_i$, can be naturally
deduced from the bias potential. % $V_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p^*_i$ is replaced by $\ln Z_i$,
  $p_i$ is the weight $\zeta_i$,
  and the bias potential $V_i$ is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

@@

The second type of methods,
which we will refer to as the adaptive FDS methods,
address the above shortcomings
by continuously and incrementally update the target distribution.
%For this reason, the second type of methods,
%referred to as the adaptive FDS methods,
%were invented.
%
Examples include
the WL algorithm and metadynamics.
%
%These methods incrementally update the bias potential
%and can approximately sample the target distribution.
%
However,
%they are less well understood theoretically, as
as mentioned, the on-the-fly updates break the microscopic reversibility
of the underlying sampling mechanism.
Further, failure to properly reduce the magnitude of
the histogram-based increments
would lead to roughness or errors in the final estimate.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods adopt a schedule to
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
barducci2008}.
%
%Thus, in late stages of a long simulation,
%the updating becomes negligible and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
%if the bias potential were converged to the exact value,
%this adaptive FDS tends to the ideal equilibrium counterpart,
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
%
For example, one can show that
if the magnitude falls according to
the optimal inverse-time prescription\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014},
the WL algorithm is equivalent to the ideal equilibrium counterpart
in improving the precision of the bias potential,
delivering the maximal efficiency
(cf. Appendix \ref{sec:equilerr}).
%the bias potential in the WL algorithm
%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%with the precision of the histogram as simulation lengthens.

@@

One can show that with a constant magnitude $\alpha(t) = \alpha > 0$,
the distribution collected from
an adaptive FDS simulation
is identical to $p_i$ in long times.
%
However, as the above updates
change $V_i(t)$ continuously,
Eq. \eqref{eq:pi_p_V} no longer holds
at all times,
and the error in $V_i(t)$
depends on the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, zhou2008, morozov2007,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.
%%
%The source of the deviation is two-fold.
%%
%First, since $V_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to reduce the magnitude $\alpha$ over time
to improve the estimate of the distribution,
$p^*_i$. %via Eq. \eqref{eq:v_def}.

@@

In the original WL papers\cite{
wang2001, *wang2001pre},
the updating magnitude $\alpha$
(or $\ln f$ there)
is controlled in stages.
%
In each stage, $\alpha$
is kept constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha$ reduced by a factor of $1/2$\cite{
wang2001, *wang2001pre}.
%
While this scheme works well for early stages,
%
it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008, morozov2009,
komura2012, *caparica2012, *caparica2014}.
%
An interesting explanation\cite{
  marsili2006, barducci2008}
comes from the interpretation where
the bias potential under the WL updating scheme,
Eq. \eqref{eq:wl_update}
as a runtime average from all data collected before time, $t$,
so that the addition of a new data point only
affects the average by a weight of $1/t$,
which is the inverse of the sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
Here we will re-derive this formula
and suggest a slight modification
whereby the origin of $t$ is shifted
according to the initial error.
%
Interestingly, we may reach the same result
from optimally planning the simulation lengths in the stages
and the reduction factor, as shown in Appendix \ref{sec:wltoinvt}.

As Eq. \eqref{eq:alpha_invt} was being developed for the WL algorithm,
several similar ways of
reducing the updating magnitude
were developed for umbrella sampling and metadynamics,
particularly, the self-healing umbrella sampling
and well-tempered metadynamics\cite{
  marsili2006, barducci2008, dickson2011, dama2014}.
%
Our aim here is to find the optimal schedule $\alpha(t)$,
for a general updating scheme.

@@

First, we need to define the error in the bias potential.
%
This is done by removing two contributions from the bias potential.
%
The first contribution comes from the target value,
Eq. \eqref{eq:Vi_target}.
%
For this purpose,
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  V_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i$'s are
equivalent to the updates to $V_i$'s,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V1},

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution $\pi_i(t)$,
%
the real deviation is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
In this way,
the variable, $x_i(t)$,
represents the actual deviation of the bias potential, $V_i(t)$,
and the total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \, \left\langle x_i^2(t) \right\rangle
  .
\label{eq:error_sum}
\end{equation}

Note that the distribution, $\pmb \rho$,
introduced above is independent of
the target distribution $\mathbf p$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is a flat distribution,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$
}%
Below, we will assume the noise is white,
such that
$$
\sigma_{ij}(t) = \Gamma_0 \, \delta(t)
$$
for some $\Gamma_0$,
where $\delta(t)$ is the Dirac's delta function.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of $\sigma_{ij}(t)$,
%
and for a non-white noise, 
the equivalent value of $\Gamma_0$ can be obtained
from the integral of the autocorrelation function
$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$.

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ r = 1 }^n
    \rho_r
    \frac{ \zeta_r(t) }
         {     p_r    }
  .
\label{eq:xi_def}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ h_r } { p_r }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} and \; \eqref{eq:h_ave}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ \langle h_r \rangle } { p_r }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:rhox_sum} \;and\; \eqref{eq:h_ave}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

The total error
at the end of the simulation, $t = T$,
as defined in Eq. \eqref{eq:error_sum},
now depends implicitly on the function $\alpha(t)$
via $x_i(t)$.
%
Our aim is to find the $\alpha(t)$
that minimizes this error.
%
It suffices to consider
a one-variable problem that
minimizes each individual
$\left\langle x_i^2(T) \right\rangle$.

>>>>

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for a class of free energy calculation methods including
  the Wang-Landau (WL) algorithm and metadynamics.
  %
  These methods rely on adaptive construction of
  a bias potential that offsets
  the potential of mean force by histogram-based adaptive updates.
  %
  The updating magnitude should decrease over time
  to reduce the noise in the bias potential,
  and one may speed up the convergence by choosing an optimal schedule
  of the updating magnitude.
  %
  %Here, we will show that the optimal schedule
  %can be derived
  %from a mechanical analogy, in which
  %the schedule corresponds to
  %the velocity of a free particle with a position-dependent mass
  %and the final error serves as the action.
  %%
  %Therefore, the optimal schedule follows from
  %the equation of motion of the particle.
  %
  We will show that
  the optimal schedule depends on the updating scheme (WL or metadynamics),
  and the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  the WL updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is the inverse number of sampling steps with a proper shift,
  as given by the $1/t$ prescription.
  %
\end{abstract}

@@

The Wang-Landau (WL) algorithm\cite{wang2001, *wang2001pre}
and metadynamics\cite{huber1994,
*laio2002, *laio2008, *barducci2011, *sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
Both techniques employ regular updating of the bias potential
to discourage future visits to the current configuration
by incrementally elevating the bias potential there.
%
The difference lies
in the updating window function,
which takes the form of a discrete
$\delta$-function (a boxcar function one histogram bin wide)
in the WL case,
or that of a Gaussian
in the metadynamics.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{zhou2005, morozov2007, zhou2008},
and for convergence
one  has to gradually decrease
the updating magnitude.

@@

In this study,
we present a method of computing
the optimal schedule
for a general adaptive FDS method,
including the WL algorithm and metadynamics.
%
Our method maps the optimization problem to a mechanic one,
in which the schedule plays the role of the velocity of
a free particle with a position-dependent mass,
and the final error becomes the action.
%
In this way, the optimal schedule
can be deduced from the equation of motion
of the particle that minimizes the action.
%
As the mass is defined from the updating window function,
the resulting optimal schedule
is different in the WL and metadynamics cases.
%
While the optimal schedule in the WL case
is given by the known inverse-time formula,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
In either case, however,
the initial updating magnitude
can be optimally set to
half of the previous equilibrium value,
which amounts to a shift of the origin of time in the WL case.
%
Further, we can generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the above origin-shifted inverse-time schedule
turns out to be optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help to better
plan and optimize FDS simulations.

@@

For a large system at temperatures of interest,
the distribution, $p^*_i$, often tends to
be localized,
%
and to explore the global shape
of the PMF, $-\ln p^*_i$,
it is often advantageous to perform
biased sampling that targets
a wider and smoother distribution, $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

@@

We wish the biased equilibrium distribution, $\pi_i$,
to be the same as the intended distribution, $p_i$,
which is usually flat,
i.e., $p_i$ is constant for all $i$'s,
or nearly flat\cite{
dayal2004, *trebst2004, *zhang2007, barducci2008, singh2011}.
%
We will broadly call such a biased simulation
a flat-distribution-sampling (FDS) simulation.
%
To achieve this distribution, the bias potential must satisfy
%
\begin{equation}
  V_i \to \ln p^*_i - \ln p_i + \mathrm{const.}
  ,
  \label{eq:Vi_target}
\end{equation}
%
according to Eq. \eqref{eq:pi_p_V1}.
%which allows us to deduce the PMF, $-\ln p^*_i$,
%directly from the bias potential.
%
This relation thus sets a practical aim for the convergence of FDS
methods---the bias potential must be adjusted to make the
observed distribution, or the normalized histogram,
match the intended one.
%
Once this aim is achieved,
the PMF, $-\ln p^*_i$, can be naturally
deduced from the bias potential. % $V_i$.
\note{
  For example, in the context of simulated tempering (Zhang and Ma, 2007),
  $p^*_i$ is replaced by $\ln Z_i$,
  $p_i$ is the weight $\zeta_i$,
  and the bias potential, $V_i$, is $\ln (\tilde Z_i / \zeta_i)$.
  Upon convergence, we have $\ln \tilde Z_i \to \ln Z_i$
  up to a constant shift.
}

@@

The second type of methods,
which we will refer to as the adaptive FDS methods,
address the above shortcomings
by continuously and incrementally updating the target distribution.
%For this reason, the second type of methods,
%referred to as the adaptive FDS methods,
%were invented.
%
Examples include
the WL algorithm and metadynamics.
%
%These methods incrementally update the bias potential
%and can approximately sample the target distribution.
%
However,
%they are less well understood theoretically, as
as mentioned, the on-the-fly updates break the microscopic reversibility
of the underlying sampling mechanism.
Further, failure to properly reduce the magnitude of
the histogram-based increments
would lead to roughness or errors in the final estimate.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods adopt a schedule to
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
barducci2008}.
%
%Thus, in late stages of a long simulation,
%the updating becomes negligible and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
%if the bias potential were converged to the exact value,
%this adaptive FDS tends to the ideal equilibrium counterpart,
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
%
For example, one can show that
if the magnitude is controlled by
the optimal inverse-time prescription\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
liang2007,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014},
the WL algorithm is equivalent to the ideal equilibrium counterpart
in improving the precision of the bias potential,
delivering the maximal efficiency
(cf. Appendix \ref{sec:equilerr}).
%the bias potential in the WL algorithm
%will evolve similar to Eq. \eqref{eq:vcorr_equil}
%with the precision of the histogram as simulation lengthens.

@@

One can show that with a constant magnitude,
$\alpha(t) = \alpha > 0$,
the distribution collected from
an adaptive FDS simulation
is identical to $p_i$ in long times.
%
However, as the above updates
change $V_i(t)$ continuously,
Eq. \eqref{eq:pi_p_V} no longer holds
at all times,
and the error in $V_i(t)$
depends on the updating magnitude\cite{
  belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
  zhou2005, morozov2007, zhou2008,
  laio2005, bussi2006, poulain2006, liang2007,
  crespo2010, *atchade2011, *fort2015}.
%%
%The source of the deviation is two-fold.
%%
%First, since $V_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to reduce the magnitude, $\alpha$, over time
to improve the estimate of the distribution,
$p^*_i$. %via Eq. \eqref{eq:v_def}.

@@

In the original WL papers\cite{
wang2001, *wang2001pre},
the updating magnitude, $\alpha$,
(or $\ln f$ there)
is controlled in stages.
%
In each stage, $\alpha$
is kept constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha$ reduced by a factor of $1/2$\cite{
wang2001, *wang2001pre}.
%
While this scheme works well for early stages,
%
it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016}.

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule, $\alpha(t)$,
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{
belardinelli2007, *belardinelli2007jcp, *belardinelli2008, *belardinelli2016,
morozov2007, zhou2008,
komura2012, *caparica2012, *caparica2014}.
%
An interesting explanation\cite{
  marsili2006, barducci2008}
comes from the interpretation of
the bias potential %under the WL updating scheme, Eq. \eqref{eq:wl_update},
as a runtime average from all data collected so far,
so that the addition of a new data point only
affects the average by a weight of $1/t$,
which is the inverse of the current sample size, $t$
(cf. Appendix \ref{sec:equilerr}).
%
Here we will re-derive this formula
and suggest a slight modification
whereby the origin of $t$ is shifted
according to the initial error.
%
Interestingly, we may reach the same result
from optimally planning the simulation lengths in the stages
and the reduction factor, as shown in Appendix \ref{sec:wltoinvt}.

As Eq. \eqref{eq:alpha_invt} was being developed for the WL algorithm,
several similar ways of
reducing the updating magnitude
were developed for umbrella sampling and metadynamics,
particularly, the self-healing umbrella sampling
and well-tempered metadynamics\cite{
  marsili2006, barducci2008, dickson2011, dama2014}.
%
Our aim here is to find the optimal schedule, $\alpha(t)$,
for a general updating scheme.

@@

First, we need to define the error in the bias potential.
%
This is done by deducting two contributions from the bias potential.
%
The first contribution comes from the target value,
Eq. \eqref{eq:Vi_target}.
%
For this purpose,
we introduce a shifted bias potential
%
\begin{equation}
  v_i(t)
  \equiv
  V_i(t)
  -
  \ln \frac { p^*_i }
            { p_i }
  ,
  \label{eq:v_def}
\end{equation}
%
which should tend to a constant of $i$
upon convergence.
Note that the updates to $v_i$'s are
equivalent to the updates to $V_i$'s,
since the shift
remains a constant during the course of simulation,
%%
%In terms of $v_i$'s, Eq. \eqref{eq:pi_p_V}
%becomes
%%
%\begin{equation}
%  \pi_i
%  =
%  \frac{                p_i \, e^{-v_i} }
%       { \sum_{j = 1}^n p_j \, e^{-v_j} }
%  \propto
%  p_i \, e^{-v_i}.
%  \label{eq:pi_p_v}
%\end{equation}
%%
%According to Eq. \eqref{eq:pi_p_V1},

Second, the bias potential following Eq. \eqref{eq:wl_update}
generally grows over time.
%
Since a uniform increase of $V_i(t)$, or, equivalently, that of $v_i(t)$,
does not affect the
resulting distribution, $\pi_i(t)$,
%
the real deviation is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$
(with $\rho_i \ge 0$ and $\sum_{i = 1}^n \rho_i = 1$).
%
In this way,
the variable, $x_i(t)$,
represents the actual deviation of the bias potential, $V_i(t)$,
and the total error is given by
%
\begin{equation}
  \Err(t)
  =
  \sum_{i = 1}^n \rho_i \, \left\langle x_i^2(t) \right\rangle
  .
\label{eq:error_sum}
\end{equation}

Note that the distribution, $\pmb \rho$,
introduced above is independent of
the target distribution, $\mathbf p$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is a flat distribution,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
From Eqs. \eqref{eq:x_def} and \eqref{eq:vbar_def},
we have
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%

@@

%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \left\langle
    \zeta_i(t) \, \zeta_j(t')
  \right\rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%
Below, we will assume the noise is white,
such that
$$
\sigma_{ij}(t) = \Gamma_0 \, \delta(t)
$$
for some $\Gamma_0$,
and $\delta(t)$ is the Dirac's delta function.
%
This approximation is valid if the simulation period, $T$,
is much longer than the autocorrelation time of $\sigma_{ij}(t)$,
%
and for a non-white noise,
the equivalent value of $\Gamma_0$ can be obtained
from the integral of the autocorrelation function,
$\int_{-\infty}^\infty \sigma_{ij}(t) \, dt$.

@@

From Eqs.
\eqref{eq:v_def}-\eqref{eq:vbar_def},
\eqref{eq:rhox_sum}-\eqref{eq:h_ave},
%\eqref{eq:v_def},
%\eqref{eq:x_def},
%\eqref{eq:vbar_def},
%\eqref{eq:rhox_sum},
%\eqref{eq:vt_diffeq},
%\eqref{eq:h_split}
%\eqref{eq:h_ave},
%\eqref{eq:dvbardt},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ r = 1 }^n
    \rho_r
    \frac{ \zeta_r(t) }
         {     p_r    }
  ,
\label{eq:xi_def}
\end{equation}
%
is also a white noise as $\zeta_i(t)$.
%
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:vbar_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ h_r } { p_r }
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:h_split} and \; \eqref{eq:h_ave}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ \langle h_r \rangle } { p_r }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:rhox_sum} \;and\; \eqref{eq:h_ave}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}
